---
title: "Dowser Tutorial"
author: "Cole Jensen"
date: "2026-05-18"
output: html_document
---

Tutorial outline:
* [Overview]
* [Getting started]
* [Define clonal groups]
* [Create germlines]
* [Build and visualize trees]
* [UCA]
* [TyCHE]

## Overview

This is a tutorial for a basic walk through of making and analyzing B cell clonal families using [10x Genomics](https://www.10xgenomics.com/products/single-cell-immune-profiling) BCR (B cell receptor) sequencing data. 

Knowledge of using R is assumed. Additionally, various preprocessing steps have been assumed to be done. Further instruction for how to do said sets can be explained in other software tutorials (AIRR-flow) or in [Immcantation tutorials](https://immcantation.readthedocs.io/en/stable/getting_started/10x_tutorial.html#assign-v-d-and-j-genes-using-igblast). 

You may also reference [this page](https://immcantation.readthedocs.io/en/stable/docker/pipelines.html) for an example pipeline script to process 10x data with Immcantation's [changeo-10x](https://bitbucket.org/kleinstein/immcantation/src/master/pipelines/changeo-10x.sh) example script.

## Getting started

We will be using a simulated dataset generated using [simble](https://github.com/hoehnlab/simble/tree/main#). You can download the data from [source](). 

#### Install and load libraries

Note that you might need to install several Bioconductor packages that are dependencies. 
```{r}
# Biocondutor dependencies first 
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

biopackages <- c("ggtree", "Biostrings", "GenomicAlignments", "IRanges", "pwalgin")
new_biopackages <- biopackages[!(biopackages %in% installed.packages())]
if (length(new_biopackages)) {BiocManager::install(new_biopackages)}
```

```{r}
# now the CRAN dependencies
packages <- c("airr", "alakazam", "data.table", "dowser", "dplyr", "ggplot2", 
              "scoper")
new_pkg <- packages[!(packages %in% installed.packages())]
if (length(new_pkg)) {install.packages(new_pkg, dependencies = TRUE)}

# call the packages
packages <- append(packages, biopackages)
for (n in seq_along(packages)) {
  suppressPackageStartupMessages(library(packages[n], character.only = TRUE))
  cat(paste0(packages[n], ": ", packageVersion(packages[n]), "\n")) # print simplified package versions
}
rm(biopackages, new_biopackages, packages, new_pkg, n)
```

For the UCA portion of this tutorial will also need a few python packages installed.
```{bash, eval=FALSE, class.source="bg-info", class.output="bg-info"}
# run this code in a command terminal, not an R terminal
pip install biopython logomaker matplotlib numpy olga pandas scipy
```

It is worth noting that sometimes the Python version you install these packages to is not the same as what R will call. Please double-check the Python executable. If you are using a virtual environment, it is possible that you have installed the packages in Python, pointed to the right executable, and still end up with an error due to the Python package not being installed. To check if the packages are installed and R can use them, you can run:
```{r, eval=FALSE}
# python3 might not be your call -- it could just be called python
system2("python3", args = c("-m", "pip", "install", "missing package name", "missing package name 2"))
```

#### Read in the data 
```{r, warning=FALSE}
library(dowser)
data <- airr::read_rearrangement(".../all_samples_airr.tsv")
```

## Data preprocessing

In normal analysis you would need to a a few preprocessing steps that we will not include here such as clonal identification, cell id standardization, and removing cells that have multiple heavy chains. Information on how to do that can be found [here](https://immcantation.readthedocs.io/en/stable/getting_started/10x_tutorial.html) and [here](https://scoper.readthedocs.io/en/stable/vignettes/Scoper-Vignette/).

## Build and visualize trees

Steps:

1.  Resolve Light Chains and Create Germlines
2.  Format Clones
3.  Tree building
4.  Reconstruct intermediate sequences
5.  Visualize trees

### Resolving Light Chains

If you have light chain data (like in our example case here) you will need to run `resolveLightChains` before running `formatClones`. In order for the trees to best use the addition of light chain information, we will need to assign `clone_subgroups`, which group cells within a clone based on the light chain V and J gene and assign a subgroup to each sequence.
```{r, warning=FALSE}
comb <- resolveLightChains(data, cell = "cell_id", nproc = 1)
```

### Create Germlines

The goal is to reconstruct the sequence of the unmutated ancestor of each clone using a reference database of known alleles ([IMGT](http://www.imgt.org)), before building B cell lineage trees. Due to the challenging nature of accurately inferring the D region and the junction region for BCR sequences, this region is masked with `N`. Note that occasionally errors are thrown for some clones - this is typical and usually results in those clones being excluded.

In the example below, we read in the IMGT germline references from our Docker container. If you're using a local installation, you can download the most up-to-date reference genome by cloning the Immcantation repository and running the script:
```{bash, eval=FALSE, class.source="bg-info", class.output="bg-info"}
# get the script
wget https://github.com/immcantation/immcantation/tree/master/scripts
# run the script
bash fetch_imgtdb.sh

# you can also copy the imgt folder from docker as well dir = "/usr/local/share/germlines/imgt/"
```

Now we can use those sequences are references when creating germlines for the data
```{r, warning=FALSE, message=FALSE}
# read in IMGT data if downloaded on your own (above)
references <- readIMGT(dir = "imgt/human/vdj")

# reconstruct germlines
comb <- createGermlines(comb, references, nproc = 1, clone = "clone_subgroup_id")
glimpse(comb)
```

### Format clones

In the rearrangement table, each row corresponds to a sequence, and each column is information about that sequence. We will create a new data structure, where each row is a clonal cluster, and each column is information about that clonal cluster. The function `formatClones` performs this processing and has options that are relevant to determine how the trees can be built and visualized. For example, `traits` determines the columns from the rearrangement data that will be included in the `clones` object, and will also be used to determine the uniqueness of the sequences, so they are not collapsed. To create H+L trees, specify `chain = "HL"`.
```{r, warning=FALSE}
clones <- dowser::formatClones(comb, chain = "HL", nproc = 1, split_light = TRUE, traits = "sample_time")
```

### Build trees

Dowser offers multiple ways to build B cell phylogenetic trees. These differ by the method used to estimate tree topology and branch lengths (e.g. maximum parsimony and maximum likelihood) and implementation (IgPhyML, PHYLIP, RAxML, or R packages `ape` and `phangorn`). For all construction methods other than `phangorn's` 'pratchet' and 'pml' methods, you will need to download and compile an executable and use the `exec` call in getTrees to point to that executable. 

For multi-partitioned trees, you need to specify that you want the construction to partition by chain type. The two methods that are currently supported do so differently. For `IgPhyML` include `partition = "hl"` in the getTrees call. For RAxML include `partition = "scaled"` in the getTrees call.

Each of these methods now supports building trees with both heavy and light chains. IgPhyML and RAxML also support multi-paritioned tree construction, paritioning on the heavy and light chain sequences seperately. 
```{r, warning=FALSE}
trees <- getTrees(clones, build = "igphyml", nproc = 1, exec = exec, partition = "hl")
```

### Reconstruct intermediate sequences

Sequences of intermediate nodes are automatically reconstructed during the tree build process. To retrieve them, first plot the node numbers for each node. The function `collapseNodes` can help clean up the tree plots. 

Get the predicted intermediate sequence at an internal node in the second largest tree (dots represent IMGT gaps):
```{r, warning=FALSE}
trees <- collapseNodes(trees)

trees <- scaleBranches(trees)
plots_all <- plotTrees(trees, node_nums = TRUE, labelsize = 6, scale = 5)[[2]] +
               geom_tippoint(aes(# simble stuff )) +
               geom_tiplab(aes(label = simble stuff), offset = 0.002)

print(plots_all)
```

```{r}
# get sequence at node 20 for the first clone_id in the trees
getNodeSeq(trees, clone = trees$clone_id[1], node = 20)
```

### Visualize trees

Regardless of how you build trees, they are visualized in the same manner with the `plotTrees` function. This will return a list of `ggplot` objects in the same order as the input object. Here, we color the tips by the `day` value because we specified that column in the `formatClones` step.

Plot all of the trees:
```{r, warning=FALSE}
plots_all <- plotTrees(trees, tips = "sample_time")
plots_all[[1]]
```

You can directly save the tree plots to PDFs using `treesToPDF`.
```{r, eval = FALSE}
# save the tree plots 
treesToPDF(plots_all, "file_path")
```

## UCA Inference

The unmuated common ancestor (UCA) is the full length inferred naive BCR sequence. Dowser now supports this inference for both single and paired chain analysis. Dowser uses a combination of tree-based statistics and VDJ combination probabiliates provided by OLGA. As part of this, you will need to provide OLGA generation probabilites (either the [default models](https://github.com/statbiophys/OLGA/tree/master/olga/default_models) or create your own). This process also currently only uses [IgPhyML](https://igphyml.readthedocs.io/en/latest/). [IgPhyML compiling information](https://igphyml.readthedocs.io/en/latest/install.html) can be found here. 
```{r, warning=FALSE}
clones_UCA <- getTreesAndUCAs(clones = clones[1,], data = comb,
                               references = references,
                               exec = "igphyml/src/igphyml",
                               model_folder = "../default_models/human_B_heavy",
                               model_folder_igk = "../default_models/human_B_kappa",
                               model_folder_igl = "../default_models/human_B_lambda",
                               nproc = 1, chain = "HL", 
                               partition = "hl", split_light = TRUE)
```

The output of `getTreesAndUCAs` is the same as the `getTrees` as well an a new column called `UCA`. `UCA` is a named list with the inferred ucas for each locus in nucleotides and amino acids as well as the IMGT gapped and ungapped versions of the UCA. 
```{r}
clones_UCA$UCA[[1]]
```

## TyCHE


