---
title: "Dowser Tutorial"
output: html_document
---

Tutorial outline: \* [Overview] \* [Getting started] \* [Define clonal groups] \* [Create germlines] \* [Build and visualize trees] \* [UCA] \* [TyCHE]

## Overview

This is a tutorial for a basic walk through of making and analyzing B cell clonal families using [10x Genomics](https://www.10xgenomics.com/products/single-cell-immune-profiling) BCR (B cell receptor) sequencing data.

Knowledge of using R is assumed. Additionally, various preprocessing steps have been assumed to be done. Further instruction for how to do said sets can be explained in other software tutorials (AIRR-flow) or in [Immcantation tutorials](https://immcantation.readthedocs.io/en/stable/getting_started/10x_tutorial.html#assign-v-d-and-j-genes-using-igblast).

You may also reference [this page](https://immcantation.readthedocs.io/en/stable/docker/pipelines.html) for an example pipeline script to process 10x data with Immcantation's [changeo-10x](https://bitbucket.org/kleinstein/immcantation/src/master/pipelines/changeo-10x.sh) example script.

## Getting started

We will be using a simulated dataset generated using [simble](https://github.com/hoehnlab/simble/tree/main#). It can be found in `data/all_samples_airr.tsv`

#### Read in the data

```{r, warning=FALSE}
library(dowser)
library(dplyr)
library(ggtree)

data <- airr::read_rearrangement("data/all_samples_airr.tsv")
data$v_germline_length <- as.numeric(data$v_germline_length)
data$d_germline_length <- as.numeric(data$d_germline_length)
data$j_germline_length <- as.numeric(data$j_germline_length)
```

## Data preprocessing

In normal analysis you would need to a a few preprocessing steps that we will not include here such as clonal identification, cell id standardization, and removing cells that have multiple heavy chains. Information on how to do that can be found [here](https://immcantation.readthedocs.io/en/stable/getting_started/10x_tutorial.html) and [here](https://scoper.readthedocs.io/en/stable/vignettes/Scoper-Vignette/).

## Build and visualize trees

Steps:

1.  Resolve Light Chains and Create Germlines
2.  Format Clones
3.  Tree building
4.  Visualize trees
5.  Reconstruct intermediate sequences

### Resolving Light Chains

If you have light chain data (like in our example case here) you will need to run `resolveLightChains` before running `formatClones`. In order for the trees to best use the addition of light chain information, we will need to assign `clone_subgroups`, which group cells within a clone based on the light chain V and J gene and assign a subgroup to each sequence.

```{r, warning=FALSE}
comb <- resolveLightChains(data, cell = "cell_id", nproc = 1)
```

### Create Germlines

The goal is to reconstruct the sequence of the unmutated ancestor of each clone using a reference database of known alleles ([IMGT](http://www.imgt.org)), before building B cell lineage trees. Due to the challenging nature of accurately inferring the D region and the junction region for BCR sequences, this region is masked with `N`. Note that occasionally errors are thrown for some clones - this is typical and usually results in those clones being excluded.

Now we can use those sequences are references when creating germlines for the data

```{r, warning=FALSE, message=FALSE}
references <- readIMGT(dir = "data/imgt/human/vdj")
comb <- createGermlines(comb, references, nproc = 1, clone = "clone_subgroup_id")
```

### Format clones

In the rearrangement table, each row corresponds to a sequence, and each column is information about that sequence. We will create a new data structure, where each row is a clonal cluster, and each column is information about that clonal cluster. The function `formatClones` performs this processing and has options that are relevant to determine how the trees can be built and visualized. For example, `traits` determines the columns from the rearrangement data that will be included in the `clones` object, and will also be used to determine the uniqueness of the sequences, so they are not collapsed. To create H+L trees, specify `chain = "HL"`.

```{r, warning=FALSE}
clones <- formatClones(comb, chain = "HL", nproc = 1, split_light = TRUE, traits = "sample_time")
```

### Build trees

Dowser offers multiple ways to build B cell phylogenetic trees. These differ by the method used to estimate tree topology and branch lengths (e.g. maximum parsimony and maximum likelihood) and implementation (IgPhyML, PHYLIP, RAxML, or R packages `ape` and `phangorn`). For all construction methods other than `phangorn's` 'pratchet' and 'pml' methods, you will need to download and compile an executable and use the `exec` call in getTrees to point to that executable.

For multi-partitioned trees, you need to specify that you want the construction to partition by chain type. The two methods that are currently supported do so differently. For `IgPhyML` include `partition = "hl"` in the getTrees call. For RAxML include `partition = "scaled"` in the getTrees call.

Each of these methods now supports building trees with both heavy and light chains. IgPhyML and RAxML also support multi-partitioned tree construction, partitioning on the heavy and light chain sequences separately.

```{r, warning=FALSE}
igphyml_exec <- "/programs/igphyml/src/igphyml"
trees <- getTrees(clones, build = "igphyml", nproc = 1, exec = igphyml_exec, partition = "hl")
```

### Plot the trees and reconstruct intermediate sequences

Regardless of how you build trees, they are visualized in the same manner with the `plotTrees` function. This will return a list of `ggplot` objects in the same order as the input object. Here, we color the tips by the `sample_time` value because we specified that column in the `formatClones` step.

```{r, warning=FALSE}
trees <- collapseNodes(trees)

trees <- scaleBranches(trees) # changes the branch lengths from genetic distance to mutations
plots_all <- plotTrees(trees, node_nums = TRUE, tips = "sample_time", scale = 10) # scale changes the bar to 10 mutations

print(plots_all)
```

Sequences of intermediate nodes are automatically reconstructed during the tree build process. To retrieve them, first plot the node numbers for each node. The function `collapseNodes` can help clean up the tree plots.

Get the predicted intermediate sequence at an internal node in the second largest tree (dots represent IMGT gaps):

```{r}
getNodeSeq(trees, clone = trees$clone_id[1], node = 24)
```

### Save the tree plots

You can directly save the tree plots to PDFs using `treesToPDF`.

```{r, eval = FALSE}
treesToPDF(plots_all, "file_path")
```

## Statistical Tests

Dowser supports a number of statistical tests to provide insights into your dataset. Some of these options are a [measurable evolution test](https://dowser.readthedocs.io/en/stable/vignettes/Measurable-Evolution/) and a [discrete traits analysis](https://dowser.readthedocs.io/en/stable/vignettes/Discrete-Trait-Vignette/. Guides on how to run these analyzes and more can be found on our [website](https://dowser.readthedocs.io/en/stable/). 

## UCA Inference

The unmuated common ancestor (UCA) is the full length inferred naive BCR sequence. Dowser now supports this inference for both single and paired chain analysis. Dowser uses a combination of tree-based statistics and VDJ combination probabilities provided by OLGA. As part of this, you will need to provide OLGA generation probabilities (either the [default models](https://github.com/statbiophys/OLGA/tree/master/olga/default_models) or create your own). This process also currently only uses [IgPhyML](https://igphyml.readthedocs.io/en/latest/). [IgPhyML compiling information](https://igphyml.readthedocs.io/en/latest/install.html) can be found here.

```{r, warning=FALSE}
clones_UCA <- getTreesAndUCAs(clones = clones, data = comb,
                               references = references,
                               exec = igphyml_exec,
                               model_folder = "data/default_models/human_B_heavy",
                               model_folder_igk = "data/default_models/human_B_kappa",
                               model_folder_igl = "data/default_models/human_B_lambda",
                               nproc = 1, chain = "HL", 
                               partition = "hl", split_light = TRUE)
```

The output of `getTreesAndUCAs` is the same as the `getTrees` as well an a new column called `UCA`. `UCA` is a named list with the inferred UCAs for each locus in nucleotides and amino acids as well as the IMGT gapped and ungapped versions of the UCA.

```{r}
clones_UCA$UCA[[1]]
```

## Questions?

## Building Time Trees

Time-resolved trees are trees where the branch lengths represent time instead of the number of mutations. This can be useful for understanding the timing of events in B cell evolution. BEAST2 is a popular tool for building time-resolved trees, and Dowser offers integration so you can run BEAST2 directly from R.

BEAST2 is an ecosystem, and there are many packages that can be useful for different analyses. For B cell analyses, the TyCHE and rootfreqs packages will allow us to specify a germline sequence and use a clock model that accounts for the heterogeneous rates of evolution across B cell subtypes.

BEAST2 is an ecosystem, and there are many packages that can be useful for different analyses. For B cell analyses, the TyCHE and rootfreqs packages will allow us to specify a germline sequence and use a clock model that accounts for the heterogenous rates of evolution across B cell subtypes.

#### Set up for BEAST2

To run BEAST2 with Dowser, you will need to specify the path to the BEAST2 executable and the path to the XML template that you want to use. The XML template will specify the model and priors that you want to use for your analysis. Default templates are available at [github.com/hoehnlab/xml-templates](https://github.com/hoehnlab/xml-templates), but you can also create your own if you want to specify different models or priors.

```{r}
# if on your local device, you'll want to change this to the path to the BEAST2 executable
# linux default: "~/beast/bin/"
# Mac default: "/Applications/BEAST\ 2.7.7/bin/"
beast <- "/opt/beast/bin/beast"
strict_template <- "data/xml-templates/StrictClock_Standard_EmpFreq.xml"
tyche_template <- "data/xml-templates/TypeLinkedExpectedOccupancy_EstTraitClockRates_EmpFreq.xml"
```

Steps:

1.  Filter data
2.  Estimate GC clock rate
3.  Build time trees with TyCHE
4.  Monitor BEAST2 runs (optional)
5.  Visualize trees and use BEAST2 outputs

## Filter data

For this analysis, we will be using the same data as before, but we will need to filter it to include only the heavy chains. The column containing the timepoint information needs to be numeric, so we convert that too.

In order to use a type-linked model, we need to specify which column in our data contains the type that we wish to associate with different clock rates. TyCHE's standard model supports a two-type case, so we use GC and non-GC B cells as our two types. In this case, we have a "location" column in our data that specifies whether a B cell is from the germinal center or from another tissue, so we use that as our type.

We need to make sure that this column is kept as a trait when we format the clones, so we include it in the `traits` argument of `formatClones`. We also need to include the time information as a trait, so we include that as well.

```{r, warning=FALSE}
heavy <- filter(data, locus == "IGH")
heavy$sample_time <- as.numeric(heavy$sample_time)

type <- "location" # the column that contains the type information for the TyCHE model, in this example, whether the B cell is from the germinal center or from another tissue

tyche_clones <- formatClones(
  heavy,
  traits = c(type, "sample_time"),
  germ   = "germline_alignment"
)
```

## Estimating the GC clock rate

The type-linked clock models implemented in TyCHE link each trait to a separate molecular clock rate which can be either fixed or estimated as a parameter. The models perform best when there is prior information about the rate of one or both populations.

If you do not have an external estimate of the clock rate for each trait, you can estimate the clock rate using a using root-to-tip regression or by fitting a BEAST2 strict clock model to GC B cells.

Here we estimate the clock rate of germinal center B cells using BEAST2 with a strict clock model.

```{r estimate-gc-clock-rate, eval=FALSE}
gc_cells = filter(heavy, location=="germinal_center")
gcf = formatClones(gc_cells, traits=c("location","sample_time"),
	germ="germline_alignment")


gctrees = getTrees(gcf, build="pml", sub_model="HKY")

plotTrees(gctrees)[[1]] + geom_tippoint(aes(color=sample_time))
```

The next step takes about 5 minutes to run, depending on your machine, as a quick demonstration. Expect results to be unconverged. To get more reliable results, set larger `mcmc_length` or `iterations`.

`getTimeTreesIterate` will run BEAST2 on each clone in parallel (here, `nproc=1`, so 1 clone at a time, since we only have one!).

```{r, eval=FALSE}
# estimate clock rate of GC B cells
# if you don't care about convergence, reduce mcmc_length
gctree = getTimeTreesIterate(gcf,
	beast=beast,
	template=strict_template,
	dir="gc_estimation",
	id="gc_strict",
	time="sample_time",
	mcmc_length=1e5,
	iterations=5,
	nproc=1, 
	CLOCK_RATE_INIT=0.001,
	KAPPA_PRIOR_M=0.67,
	KAPPA_PRIOR_S=0.2,
	ignore=c("freqParameter"))
```

`getTimeTreesIterate` is designed to run each analysis iteratively, checking for convergence after each iteration. If the analyses converge before reaching the max iterations, it will stop early. It will run each analysis for `mcmc_length` MCMC samples (here, `1e5`), and it will repeat this up to `iterations` times (here, `5`), so here we have a maximum of 5e5 MCMC samples, which is far too few to expect convergence.

The convergence check is based on the ESS of the parameters reported in the log files. You can exclude parameters from this ESS check using the `ignore` argument (here, we ignore `freqParameter`, as it is a fixed value).

The effective sample size (ESS) of a parameter is a measure of how much independent information your MCMC sample contains. Even though MCMC generates many samples, they are typically autocorrelated—each sample depends on the previous one. ESS is the number of independent draws from the target distribution with the same estimation power and can be thought of as the sample size for that parameter.

A higher ESS means your sample more reliably represents the posterior distribution. Low ESS indicates that you need more information, which can indicate the need for longer MCMC runs. We typically recommend an ESS of at least 200 for each estimated parameter.

We can use results we ran ahead of time to look at the difference in outputs.

```{r, eval=FALSE}
gctree_converged <- readRDS("data/gctree_converged.RDS")
print(gctree$parameters[[1]] %>% filter(item=="geneticClockRate"))
print(gctree_converged$parameters[[1]] %>% filter(item=="geneticClockRate"))
```

While the mean clock rate happens to be similar between the unconverged and converged analyses, the ESS of the clock rate parameter is much higher in the converged analysis, so we can be more confident in that estimate.

To get the mean clock rate across the trees, we can filter the parameters table for `geneticClockRate` and take the mean of the `mean` column. This will give us an estimate of the clock rate for GC B cells that we can use in our TyCHE model.

```{r, eval=FALSE}
gcrate_tree = mean(sapply(gctree_converged$parameters, function(x)filter(x,item=="geneticClockRate")$mean))
print(gcrate_tree)
```

If it is not feasible to run a strict clock analysis, you can use the slope from a root-to-tip regression. Here, we estimate the clock rate of germinal center B cells using a root-to-tip regression.

```{r, eval=FALSE}
gcrate_slope = mean(correlationTest(gctrees, time="sample_time")$slope)
print(gcrate_slope)
```

## Build time trees with TyCHE

We can now run a type-linked TyCHE model using the estimated GC rate. Here, we use the *TypeLinkedExpectedOccupancy* model, which uses an expected occupancy method to determine the proportion of each branch in each state.

Features of this template:

-   Allows estimation of clock rates:
    -   we provide values of the mean (`TRAIT_RATE_MEAN_1`, `TRAIT_RATE_MEAN_2`) and standard deviation (`TRAIT_RATE_SIGMA_1`, `TRAIT_RATE_SIGMA_2`) for the prior normal distributions of each clock rate.
-   Uses empirical nucleotide frequencies as the equilibrium frequencies.
    -   Dowser will automatically calculate these frequencies from the input sequences.
    -   Recommended for most data, especially BCRs.

The next step takes about 10 minutes to run, depending on your machine, as a quick demonstration. Expect results to be unconverged. To get more reliable results, set larger `mcmc_length` or `iterations`.

```{r, eval=FALSE}
mixed_trees <- getTimeTreesIterate(
  tyche_clones,
  beast    = beast,
  template = tyche_template,
  trait    = "location", # this should be just one trait column
  time     = "sample_time",
  dir      = "tyche_estimation",
  id       = "tyche_tutorial",
  log_every = "auto",
  nproc     = 1,
  KAPPA_PRIOR_M = 0.67, # log normal prior, corresponds to a mean of about 2
  KAPPA_PRIOR_S = 0.2,
  TRAIT_RATE_MEAN_1 = gcrate_tree,
  TRAIT_RATE_MEAN_2 = 0.000001,
  TRAIT_RATE_SIGMA_1 = gcrate_tree * 0.01,
  TRAIT_RATE_SIGMA_2 = 0.001,
  RATE_INDICATORS = "1 0",
  TYPE_SWITCH_INIT = 0.005,
  TYPE_SWITCH_ALPHA = 0.001,
  TYPE_SWITCH_BETA = 5.0,
  TRANSITION_RATE_1_INIT = 1.0,
  TRANSITION_RATE_2_INIT = 1.0,
  TRANSITION_RATE_ALPHA_1 = 0.1,
  TRANSITION_RATE_BETA_1  = 1.0,
  TRANSITION_RATE_ALPHA_2 = 0.1,
  TRANSITION_RATE_BETA_2  = 1.0,
  log_target   = 2000,
  mcmc_length  = 1e6,
  ignore       = c("freqParameter"),
  iterations   = 4
)
```

`getTimeTreesIterate` will run BEAST2 on each clone in parallel (here, `nproc=1`, so 1 at a time).

To capture sufficient information about the posterior distribution while keeping log files from becoming overly large or unwieldy, we provide the option to set `log_every="auto"`. This will automatically set the logging frequency based on the `mcmc_length` and `log_target` (here, `2000`, so we aim to have around 2000 samples in the log file). You can also set a fixed logging frequency by providing an integer value.

The rate indicators (`RATE_INDICATORS`) specify which traits can transition to each other. In a primary immune response we recommend setting this to `"1 0"`, as GC B cells can transition to other tissues, but not vice versa. If your data comprises chronic infections or repeated vaccinations, you may want to allow transitions in both directions, so you would set this to `"1 1"`. Note: traits are sorted ASCII alphabetically.

You can also specify alpha (shape) and beta (rate) values for the prior gamma distributions of the transition rates between traits. We recommend setting the same prior for each transition rate except in rare cases.

The prior distribution on kappa is used by the nucleotide substitution model, and we recommend these values for BCR analyses.

For more details or future reference, see `?getTimeTreesIterate`.

## Monitor BEAST2 runs (optional)

You can monitor the progress of BEAST2 runs using tools like Tracer and FigTree. Let's step away from R and take a look at those.

## Visualize the results and use BEAST2 parameter outputs

After the analyses have converged, you can visualize the time trees.

Since the analysis above takes some time to run, we will load in the results from a previous run for the rest of this tutorial.

```{r load-mixed-trees, echo=FALSE}
mixed_trees <- readRDS("data/mixed_trees.RDS")
```

Note: plotTrees sets a default value for the scale bar of 0.01, which is appropriate for trees with genetic distance branch lengths (mutations per site), but time trees typically require a larger scale bar. In this case, we know the data spans 100 days, so we set `scale=100` to make the scale bar more visually interpretable.

```{r plot-mixed-trees-result}
plotTrees(mixed_trees, scale=100, nodes=TRUE, tips="location")[[1]] + geom_point(aes(fill=location), pch=21, size=3)
```
If we want to visualize how much of each branch is estimated in each state, we can plot the "expected occupancy" of each branch. This shows how
much time TyCHE estimates each branch spends in the GC. The expected occupancy is a value between 0 and 1, so we can use a gradient color scale to visualize it.

```{r plot-mixed-trees-eo}
colors = RColorBrewer::brewer.pal(3,"Set1")
names(colors) = c("GC", "Other", "Ambig.")
eo_line_color_scale <- scale_color_gradient(
    low = colors["Other"],
    high = colors["GC"],
    breaks = c(0, 0.25, 0.5, 0.75, 1),
    limits = c(0, 1)
)

ggtree(mixed_trees$trees[[1]], aes(color = as.numeric(expectedOccupancies)),linewidth = 0.75) +
  geom_point(aes(fill = location), pch = 21, size = 3, color = "black") +
  eo_line_color_scale
```
The `parameters` column of `mixed_trees` contains a table that collates the output from the BEAST2 analysis. Columns include the parameter (item), the posterior mean, standard error, standard deviation, median, 95% highest posterior density interval, autocorrelation time (ACT), effective sample size, and geometric mean.

The effective sample size (ESS) of a parameter is a measure of how much independent information your MCMC sample contains. Even though MCMC generates many samples, they are typically autocorrelated—each sample depends on the previous one. ESS is the number of independent draws from the target distribution with the same estimation power and can be thought of as the sample size for that parameter.

A higher ESS means your sample more reliably represents the posterior distribution. Low ESS indicates that you need more information, which can indicate the need for longer MCMC runs. We typically recommend an ESS of at least 200 for each estimated parameter.

The autocorrelation time (ACT) of a parameter measures how strongly each sample in the MCMC chain depends on previous samples. The ACT tells you how long the MCMC chain takes to produce a roughly independent sample. ACT is inversely related to ESS.

If we're interested in the estimated tree height, we can filter the parameters table for `TreeHeight`:

```{r tree-height-parameter}
print(mixed_trees$parameters[[1]] %>% filter(item=="TreeHeight"))
```

Our model likely hasn't converged, with multiple parameters having ESS values below 200, particularly the posterior which describes how well the model has converged as a whole. The mean tree height is around 236 time units, with a 95% highest posterior density interval from about 200 to 275 time units. Since we know the data spans 200 time units, this is a high estimate, but this is unsurprising given that the ESS is below 200 and the analysis has likely not converged.

The parameters available will depend on the model you used and what is specified for logging in the XML template. In this case, we can see all the items that were logged:

```{r logged-parameters}
print(mixed_trees$parameters[[1]]$item)
```

These include the posterior, likelihood, and prior probabilities of the full model; the tree likelihood; estimated values of the tree height, the clock rates for each trait (`typeLinkedRates`), the relative transition rates between traits (`relativeGeoRates`), the rate of switching traits (`typeSwitchClockRate`); parameters relating to BayesianSkyline (`BayesianSkyline`, `bPopSizes`, `bGroupSizes`); and some fixed parameters that are included in logging for record-keeping convenience (the kappa value of the HKY substitution model, the empirical frequencies of the nucleotides, the frequencies of the traits).

If you want to revisit an analysis and no longer have the `mixed_trees` object in your R environment, you can use `readBEAST` to read in the BEAST log and tree files from the directory (`dir`) you specified in `getTimeTreesIterate`. Because of this, it is important to always specify a unique combination of `dir` and `id` for each analysis.

```{r read-beast-output, eval=FALSE}
mixed_trees <- readBEAST(tyche_clones, dir="tyche_estimation", id="tyche_tutorial", beast=beast, trait=trait)
```

See `?readBEAST` for more details.

You can find all of BEAST's output files, including the trees, logs, console logs, and TreeAnnotator outputs, in the `dir` you specified (here, "temp"). You can view these files using BEAST tools such as Tracer.

## Additional configuration examples

We assumed that the example above was a primary GC. TyCHE also works in cases where you expect recall GCs or other GC re-entry. To see examples of other scenarios and how you might change your parameters to fit those scenarios, check out [other files]() <!-- TODO: add these -->
